Top-down versus bottom-up cohesiveness in microbial community coalescence

### 2021-05-13 draft/draft.tex
updated affiliations (but check with co-authors)
new figure 2 (incorporated former figure S4), added figure 3, notes in the draft for later editing

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness
in microbial community coalescence}
\author[1,2,*]{Juan Diaz-Colunga}
\author[1,2,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4,5]{Mikhail Tikhonov}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Cell Metabolism, Agios Pharmaceuticals, Inc.,
Cambridge, MA, USA}
\affil[4]{Department of Applied Physics,
Stanford University, Stanford, CA, USA}
\affil[5]{Department of Physics,
Washington University, St. Louis, MO, USA}

@@

The phenomenon by which entire microbiomes invade one another has been termed
\textit{community coalescence} \cite{Rillig2015}.
Ecologists have long contemplated the idea that interactions between multiple co-invading species
can produce correlated invasional outcomes
\cite{Gilpin1994,Simberloff1999,Grosholz2005,Simberloff2006,Gurevitch2006,Green2011,
Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
\iffalse
One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
positive interactions between co-invading species can enhance their invasive success and
facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
\fi
However, and
in spite of its clear potential importance, the role of coalescence in microbiome assembly is
only beginning to be addressed and little is known about the mechanisms that govern it and its
potential implications.
Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
as well as more recent work
\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
suggest that high-order invasion effects are common
during community coalescence. Communities that have a previous history of coexistence may exhibit an
emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
same community \cite{Livingston2013,Sierocinski2017}.
The situation where ecological partners in the invading community recruit each other into the final
coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.

@@

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then performed all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}),
and we quantified coalescence outcomes using the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yielded similar results,
with weaker effects when the metric used
accounted only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
(\figref{figS4}, $R^2=0.04$, $p>0.05$).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 (\figref{figS4}, $R^2=0.83$, $p<10^{-8}$).

@@

To better understand the mechanisms of ecological co-selection in our experiments,
we ran a set of simulations of community coalescence. We used a microbial consumer-resource
model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks \cite{Goldford2018}, for which the
microCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we replicated our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were then allowed to stabilize through 20
growth-dilution cycles. Stable communities were then used to reproduce our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

We carried out two sets of simulations under two different regimes.
In one case, we generated a library of 30 resources divided into 3 resource classes
(\methodsref{sim}). This corresponds to a situation where the digestion of a resource
can yield a wide variety of secreted byproducts.
We refer to this scenario as the \textit{complex resource} regime.
Our second set of simulations was performed under a \textit{simple resource} regime, where we
generated a library of only 15 resources belonging to a same class.
For the simple resource, we found that coalescence outcomes were highly correlated with
the competitive ability of the community dominants in pairwise competition
(\figref[c]{fig2}),
indicating a large degree of top-down positive co-selection.
These dominants were able to reach similar fractions in the final communities when invading alone
with respect to when invading with their cohorts
(\figref[d]{fig2}),
suggesting that bottom-up co-selection is weak in this scenario.
On the other hand,
simulations under the complex resource regime evidenced weak top-down co-selection but 
common and strong bottom-up recruitment (\figref[e-f]{fig2}), with negative co-selection remaining rare.
Simulations suggest that community cohesiveness is dependent on the metabolic interactions among species,
which is in line with previous findings \cite{Goldford2018, Estrela2020}.
These interactions are determined by the metabolic profile of the primary resource in the conditions
of our simulations and, in turn, modulate the strength and direction of ecological co-selection.

@@

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=11cm,keepaspectratio]{figs/fig2.pdf}
\caption{\textbf{Co-selection in microbial community coalescence.}
\textbf{a.}~Coalescence outcomes are quantified by the relative Bray-Curtis similarity
($Q$) between the coalesced and invasive communities.
These outcomes are predicted by the pairwise competition between the invasive and
resident dominant species
($R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
A higher correlation is consistent with a scenario of
top-down positive co-selection where dominants recruit their cohorts for the final
coalesced community. Two biological replicates per experiment are plotted individually.
\textbf{b.}~We represent the frequency reached by
the invasive dominant species when they invade the resident communities on their own versus
when they are in the company of their cohort.
Three scenarios are possible: green and red
shaded areas represent limit cases of positive (recruitment) or negative (antagonism)
bottom-up co-selection, gray area corresponds to situations where invasive dominant
species can invade with equal success regardless of the presence of their cohorts.
Data shows that positive co-selection is common, whereas antagonistic
co-selection is rare in our experiments.
Two biological replicates per experiment are plotted individually.
\textbf{c-d.} Simulations of community coalescence with a consumer-resource model
and a metabolically simple supplied resource yield a strong degree of top-down positive
co-selection
($R^2=0.81$, $p<10^{-15}$)
and few instances of bottom-up co-selection.
\textbf{e-f.} Simulations with a metabolically complex resource result in weaker
top-down co-selection
($R^2=0.22$, $p<10^{-5}$)
but frequent bottom-up positive co-selection.}
\label{fig2}
\end{figure}

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=8cm,keepaspectratio]{figs/figS4.pdf}
\caption{\textbf{Pairwise competition of dominants with or without their cohorts.}
In the horizontal axis, we plot the relative frequency of the invasive dominant species in head-to-head
pairwise competition. In the vertical axis, we plot the same relative frequency when competition takes
place in the presence of the cohorts, i.e. during community coalescence.
$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.}
\label{figS4}
\end{figure}

>>>>

% title, authors, affiliations
\title{Top-down versus bottom-up cohesiveness
in microbial community coalescence}
\author[1,2,*]{Juan Diaz-Colunga}
\author[1,2,*]{Nanxi Lu}
\author[1,2,*]{Alicia Sanchez-Gorostiaga}
\author[3]{Joshua E. Goldford}
\author[4]{Mikhail Tikhonov}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Cell Metabolism, Agios Pharmaceuticals, Inc.,
Cambridge, MA, USA}
\affil[4]{Department of Physics,
Center for Science \& Engineering of Living Systems,
Washington University in St. Louis, St. Louis, MO, USA}

@@

The phenomenon by which entire microbiomes invade one another has been termed
\textit{community coalescence} \cite{Rillig2015}.
Ecologists have long contemplated the idea that interactions between multiple co-invading species
can produce correlated invasional outcomes
\cite{Gilpin1994,Simberloff1999,Grosholz2005,Simberloff2006,Gurevitch2006,Green2011,
Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
However, and
in spite of its clear potential importance, the role of coalescence in microbiome assembly is
only beginning to be addressed and little is known about the mechanisms that govern it and its
potential implications.
Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
as well as more recent work
\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
suggest that high-order invasion effects are common
during community coalescence. Communities that have a previous history of coexistence may exhibit an
emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
same community \cite{Livingston2013,Sierocinski2017}.
The situation where ecological partners in the invading community recruit each other into the final
coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.

@@

If communities being coalesced were highly cohesive from the top-down,
the dominant species would co-select the rarer members of its community during coalescence
(\figref[e]{fig1}, left panels).
In this scenario, we would expect the outcome of community coalescence to be predicted by
which of the two dominants is most competitive in pairwise competition.
Analogously, competition between dominants should be affected only weakly by the presence or
absence of the cohorts, that would play a passive role under these conditions.
To test this hypothesis, we performed all pairwise competitions between dominant species
in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
the cultures for seven serial transfers, roughly 42 generations
(\methodsref{competitions}).
We then performed all possible pairwise community coalescence experiments by mixing
equal volumes of the communities and propagating the resulting cultures for seven extra
transfers
(\figref[f]{fig1}),
and we quantified coalescence outcomes using the similarity between the coalesced and the
invasive communities
(\methodsref{metrics}).
The frequencies of all species in both community-community and dominant-dominant competitions
were determined by 16S Illumina sequencing
(\methodsref{sequencing}).

\mr Rearrange this section to follow the flow of the new figure 2.
Also introduce the model here and cook a nice experiment-theory-experiment sandwich later.

We found that, for communities assembled in the glutamine environment,
the pairwise competitive ability of an invasive dominant
is only weakly predictive of the performance of the invasive community in coalescence
(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
but this correlation was stronger for the citrate communities
(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
Alternative quantifications of community distance yielded similar results,
with weaker effects when the metric used
accounted only for the presence/absence of specific species and not for their relative abundance in
the communities
(\figref{figS2}).
All these metrics include the presence of the dominant species themselves. To better disentangle the effect
that these dominants have on the other members of their communities, we repeated the analysis this time
excluding the dominant species from the compositional data, finding that our results still hold
(\figref{figS3}).
Furthermore, we observed that, in the glutamine communities,
the relative frequency of a dominant against another in head-to-head pairwise competition
is barely predictive of its relative frequency against that same other dominant when the
cohorts are present too, i.e. after coalescence
(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
On the other hand, pairwise competition between dominants of the citrate communities
is highly correlated with the relative abundance of said dominants in community coalescence
 (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).

@@

To better understand the mechanisms of ecological co-selection in our experiments,
we ran a set of simulations of community coalescence. We used a microbial consumer-resource
model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
for Python \cite{Marsland2020}. We chose this modeling framework because
communities assembled under our experimental conditions (natural microbiomes re-assembled into
multispecies communities through serial growth-dilution cycles in synthetic minimal media
with a single carbon source)
have been shown to be
sustained by dense metabolic cross-feeding networks
\cite{Goldford2018,Estrela2020},
for which the microCRM provides a good description.
We implemented a new feature in the package that makes individual species
able to secrete different sets of byproducts when consuming a same resource
(\methodsref{sim}).
We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
types of byproducts when consuming a resource. These assumptions are supported by the experimental
observation that individual species can indeed secrete different sets of few metabolites
to the environment when consuming a primary resource
\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
We reason that these choices are necessary to potentially generate community cohesiveness:
if the secretions of all species were roughly similar, the effect of interchanging members of
a community would be minimal. But in practice, species with a history of coexistence make up
cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
In short,
we reproduced our experimental protocol \textit{in silico}
by generating a library of species that we divided into two pools. Each pool was used to seed a
collection of invasive and resident communities, that were allowed to stabilize through 20
growth-dilution cycles. We then mixed pairs of these stable communities to carry out our coalescence
and dominant-dominant competition experiments,
as well as the invasion experiments from the dominants alone.
A full description of our modeling framework can be found in the \methodsref{sim} section.

\iffalse
We carried out two sets of simulations under two different regimes.
In one case, we generated a library of 30 resources divided into 3 resource classes
(\methodsref{sim}). This corresponds to a situation where the digestion of a resource
can yield a wide variety of secreted byproducts.
We refer to this scenario as the \textit{complex resource} regime.
Our second set of simulations was performed under a \textit{simple resource} regime, where we
generated a library of only 15 resources belonging to a same class.
For the simple resource, we found that coalescence outcomes were highly correlated with
the competitive ability of the community dominants in pairwise competition
(\figref[c]{fig2}),
indicating a large degree of top-down positive co-selection.
These dominants were able to reach similar fractions in the final communities when invading alone
with respect to when invading with their cohorts
(\figref[d]{fig2}),
suggesting that bottom-up co-selection is weak in this scenario.
On the other hand,
simulations under the complex resource regime evidenced weak top-down co-selection but 
common and strong bottom-up recruitment (\figref[e-f]{fig2}), with negative co-selection remaining rare.
Simulations suggest that community cohesiveness is dependent on the metabolic interactions among species,
which is in line with previous findings \cite{Goldford2018, Estrela2020}.
These interactions are determined by the metabolic profile of the primary resource in the conditions
of our simulations and, in turn, modulate the strength and direction of ecological co-selection.
\fi

\mr Reframe the entire simulations section (include it in previous two sections?).
Sandwich: the model predicts a) that positive bottom-up co-selection is much more common than
negative bottom-up co-selection... and we see it in the experiments; and b) that in the instances
where bottom-up co-selection is strong (green area in figure 3...), the relative frequency of the dominants
in head-to-head pairwise competition is a very poor predictor of community coalescence outcomes...
which we confirm experimentally.
Conclusions: results and previous theoretical and experimental work on different systems (and on the
general relationship between metabolic interactions and community structure) suggest that...
the strength and direction of ecological co-selection is determined by the underlying metabolic networks that
shape the structure and interactions of communities assembled in synthetic minimal environments
during coalescence.
These networks are modulated by the supplied primary resource in our conditions.
When top-down co-selection is weak, bottom-up positive co-selection can be very strong, while negative
co-selection remains rare in our experiments.

@@

\mr Update with new simulation protocol.

@@

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=15cm,keepaspectratio]{figs/fig2_v3.pdf}
\caption{\textbf{Top-down co-selection in microbial community coalescence.}
\textbf{a.} Pairwise competition of dominants with or without their cohorts.
In the horizontal axis, we plot the frequency of the invasive dominant species
in head-to-head pairwise competition with the resident dominant.
In the vertical axis, we plot the same relative frequency when the two species compete
in the presence of their cohorts, i.e. during community coalescence.
$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.
\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
($Q$) between the coalesced and invasive communities.
These outcomes are predicted by the pairwise competition between the invasive and
resident dominant species.
Left panel (red): glutamine communities, $R^2=0.15$, $p<0.05$.
Middle panel (blue): citrate communities, $R^2=0.57$, $p<10^{-4}$.
A high correlation is consistent with a scenario of
strong top-down positive co-selection where dominants recruit their cohorts for the final
coalesced community. Two biological replicates per experiment are plotted individually.
Right panel (black): simulations with a microbial consumer-resource model are able
to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
\label{fig2}
\end{figure}

\begin{figure}[!h]
\centering
\internallinenumbers
\includegraphics[width=12cm,keepaspectratio]{figs/fig3.pdf}
\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
\textbf{a.} Blablabla.
\textbf{b.} Blablabla.
\textbf{c.} Blablabla.
\textbf{d.} Blablabla.
\textbf{e.} Blablabla.
\textbf{f.} Blablabla.}
\label{fig3}
\end{figure}

@@
